---
header-includes: 
  - \input{preamble.tex}
latex_engine: xelatex
fontsize: 9.5pt
output:
  pdf_document:
    fig_caption: yes
    number_sections: yes
sansfont: Lato
font: Lato
geometry: left=0.65in,right=0.65in,top=0.35in,bottom=0.5in
urlcolor: null
---

\raggedright

```{r, include=FALSE}
options(tinytex.verbose = TRUE)
```

```{r rmarkdown-setup, include=FALSE}
library(data.table)
library(flextable)
library(ggthemes)
library(glue)
library(gt)
library(janitor)
library(lubridate)
library(plyr)
library(scales)
library(tidyverse)
library(zoo)
```

# general functions

```{r include = FALSE}
# label/table functions
percentLabel <- function(x, digits = 2, format = "f", ...) {
  paste0(formatC(x, format = format, digits = digits, ...), "%")
}

dollarLabel <- function(x, ...) {
  paste0("$", comma(x), ...)
}

# get current date
month_day <- format(Sys.time(), "%m%d")
month_day
```

# lending club - general cleaning function

```{r}
# cleaning function
lending_club_clean = function(data, ...) {
  data %>% 
    # relabel 'loan status' var 
    dplyr::mutate(loan_status2 = factor(case_when(str_detect(loan_status, "Late") ~ "Late",
                                                  TRUE ~ as.character(loan_status)),
                                        levels = c("Fully Paid", "Current", "Late",
                                                   "Charged Off", "In Grace Period", "Default")),
                  # relabel `grade` var
                  grade = factor(case_when(grade == "F" ~ "FG",
                                           grade == "G" ~ "FG",
                                           TRUE ~ as.character(grade)),
                                 levels = c("A", "B", "C", "D", "E", "FG")),
                  # create numeric 'interest rate' var
                  int_rate_dbl = as.double(str_trim(str_replace_all(int_rate, "%", "")))/100) 
}
```

# lending club table reproducability functions

```{r}
# part 1
# calculate total amount issued by loan status
# (for data with multiple quarters or at least a dataframe with a column indicating which quarter)
part1_total_issued = function(data, ...) {
  data %>% 
    dplyr::select(quarter, grade, loan_status2, loan_amnt) %>%
    dplyr::group_by(quarter, grade, loan_status2) %>%
    # calculate loan amount issued by quarter, grade, and loan status
    dplyr::summarise(issued = sum(loan_amnt)) %>%
    dplyr::group_by(quarter, grade) %>%
    # calculate total loan amount issued by quarter and grade only
    dplyr::mutate(`Total Issued` = sum(issued)) %>%
     # reshape dataframe long to wide
    pivot_wider(names_from = loan_status2, 
                values_from = c(issued)) %>%
    bind_rows(dplyr::group_by(., quarter) %>%
              dplyr::summarise(across(where(is.numeric), ~ sum(.x, na.rm = TRUE))) %>%
    dplyr::mutate(grade = "All")) %>%
    dplyr::mutate(across(where(is.numeric), ~ dollarLabel(.x)),
                  grade = factor(grade, 
                                 levels = c("A", "B", "C", "D", "E", "FG", "All"))) %>%
    dplyr::arrange(quarter) %>%
    dplyr::select(quarter, grade, `Total Issued`, `Fully Paid`, `Current`, `Late`, `Charged Off`) %>%
    na_if("$NA")
}  
```

```{r}
# part 2
# calculate principal payments received
# (for data with multiple quarters or at least a dataframe with a column indicating which quarter)
part2_principal_calc = function(data, ...) {
  data %>% 
    dplyr::group_by(quarter, grade) %>%
    # calculate total principal payments received
    dplyr::summarise(`Principal Payments Received` = sum(total_rec_prncp)) %>%
    dplyr::ungroup() %>%
    # calculate total row for each grade by quarter
    bind_rows(group_by(., quarter) %>%
              summarise(`Principal Payments Received` = sum(`Principal Payments Received`)) %>%
    dplyr::mutate(grade = "All")) %>%
    dplyr::mutate(`Principal Payments Received` = dollarLabel(`Principal Payments Received`),
                  grade = factor(grade, 
                                 levels = c("A", "B", "C", "D", "E", "FG", "All"))) %>%
    dplyr::arrange(quarter)
}
```

```{r}
# part 3
# calculate interest payments received
# (for data with multiple quarters or at least a dataframe with a column indicating which quarter)
part3_rec_int = function(data, ...) {
  data %>% 
    dplyr::group_by(quarter, grade) %>%
    # calculate total principal payments received
    dplyr::summarise(`Interest Payments Received` = sum(total_rec_int)) %>%
    dplyr::ungroup() %>%
    # calculate total row for each grade by quarter
    bind_rows(group_by(., quarter) %>%
              summarise(`Interest Payments Received` = sum(`Interest Payments Received`)) %>%
              mutate(grade = "All")) %>%
    dplyr::mutate(`Interest Payments Received` = dollarLabel(`Interest Payments Received`),
                  grade = factor(grade, 
                                 levels = c("A", "B", "C", "D", "E", "FG", "All"))) %>%
    dplyr::arrange(quarter)
}
```

```{r}
# part 4
# calculate average interest rate
# no function created as averages for all and by group were calculated separately, then joined
```

```{r}
# part 5
# calculate adj. net annualized return (in progress)
# reference: https://www.lendingclub.com/public/lendersPerformanceHelp.action
```

## data import

```{r}
# setwd("~/Documents/GitHub/challenge")

# pinpoint all zipped files in raw data folder
temp_unzip <- list.files(path = "~/Documents/GitHub/challenge/zipped data", pattern = "*.zip", full.names = TRUE)

# unzip all files and store in clean data folder
ldply(.data = temp_unzip, .fun = unzip, exdir = "~/Documents/GitHub/challenge/clean data")
```

```{r}
# setwd("~/Documents/GitHub/challenge")

# import all unzipped .csv dataframes in folder
lending_club <- list.files(path = "~/Documents/GitHub/challenge/clean data",  
                       pattern = "*.csv", full.names = TRUE) %>% 
  lapply(fread) %>%                             
  bind_rows() %>%
  # create `quarter` var based on `issue_d` var for date loan issued
  separate(issue_d, into = paste("issued", 1:2, sep = "_"), sep = "-") %>%
  dplyr::mutate(quarter_month = factor(case_when(str_detect(issued_1, c("Jan|Feb|Mar")) ~ "Q1",
                                                 str_detect(issued_1, c("Apr|May|Jun")) ~ "Q2",
                                                 str_detect(issued_1, c("Jul|Aug|Sep")) ~ "Q3",
                                                 str_detect(issued_1, c("Oct|Nov|Dec")) ~ "Q4"),
                                       levels = c("Q1","Q2","Q3","Q4")),
  # reorder levels in 'quarter' var
  # this will vary depending on timeframe of future datasets, so 
  # it is kept separate from the cleaning function
                quarter = factor(paste(issued_2, quarter_month, sep = "_"),
                                 levels = c("2018_Q3","2018_Q4","2019_Q1"))) %>%
  # perform cleaning function
  lending_club_clean() 

glimpse(lending_club)
```

## exploratory data analysis

```{r}
# quality check / exploratory analysis
# check count of all applications by grade and by quarter
lending_club_grade_count <- lending_club %>%
  dplyr::count(quarter, grade) %>%
  ggplot() +
  geom_col(aes(y = grade, x = n, fill = grade)) +
  geom_label(aes(y = grade, x = n, group = quarter, label = n), 
            size = 2, color = "#000000") +
  scale_fill_viridis_d() +
  scale_x_continuous(labels = scales::comma_format()) +
  labs(title = "Number of Loan Applications,\nby Vintage Quarter and Grade", 
       fill = "Grade",
       caption = "Source: LendingClub") +
  facet_grid(cols = vars(quarter)) +
  scale_y_discrete(limits = rev) +
  theme_fivethirtyeight()

ggsave(glue::glue("output/fig1_lending_club_grade_count_{month_day}.png"), lending_club_grade_count, 
       width = 7, height = 4, device = "png")

lending_club_grade_count
```

## it seems that relative to 2018_Q3, there are significantly fewer applications in other quarters. 
## we should connect with lendingclub to confirm that they shared all the applications and didn't
## filter out any observations by accident.

```{r}
# quality check / exploratory analysis
# check count of all applications by grade and by quarter
lending_club_rate_dist <- lending_club %>%
  ggplot() +
  geom_boxplot(aes(y = grade, x = int_rate_dbl, fill = grade),
               alpha = 0.75) +
  scale_fill_viridis_d() +
  scale_x_continuous(labels = scales::percent_format()) +
  labs(title = "Interest Rate Distribution,\nby Vintage Quarter and Grade", 
       fill = "Grade",
       caption = "Source: LendingClub") +
  facet_grid(cols = vars(quarter)) +
  scale_y_discrete(limits = rev) +
  theme_fivethirtyeight()

ggsave(glue::glue("output/fig2_lending_club_rate_dist_{month_day}.png"), lending_club_rate_dist, 
       width = 7, height = 4, device = "png")

lending_club_rate_dist
```

## while there are significantly fewer observations for the two most recent quarters, the distribution  
## of interest rates seems pretty similar across quarters for each grade. though, the smaller 
## sample sizes in the two recent quarters is likely resulting in some skewed distributions, as in
## the case of grade "E" in 2018_Q4, for example. moreover, the outliers in the earliest available
## quarter also warrant further inquiry. 

```{r}
# quality check / exploratory analysis
# check total funding of all applications by loan status, grade and by quarter
lending_club_table_plot <- lending_club %>%
  dplyr::filter(loan_status2 == "Fully Paid" | loan_status2 == "Current" | 
                  loan_status2 == "Late" | loan_status2 == "Charged Off") %>%
  dplyr::group_by(quarter, loan_status2, grade) %>%
  dplyr::summarise(total = sum(loan_amnt)) %>%
  ggplot() +
  geom_col(aes(y = grade, x = total, fill = grade),
               alpha = 0.75) +
  scale_fill_viridis_d() +
  scale_x_continuous(labels = scales::dollar_format()) +
  labs(title = "Loan Amount Issued,\nby Vintage Quarter and Grade", 
       fill = "Grade",
       caption = "Source: LendingClub") +
  facet_grid(rows = vars(quarter),
             cols = vars(loan_status2)) +
  scale_y_discrete(limits = rev) +
  theme_fivethirtyeight()

ggsave(glue::glue("output/fig3_lending_club_table_plot_{month_day}.png"), lending_club_table_plot, 
       width = 11, height = 7, device = "png")

lending_club_table_plot
```

## the low loan amount issued for the recent quarters also raises a red flag, based on the amount
## issued for the first quarter. 

## data wrangling steps 1-6

## step 1
## total issued by loan status*
## *missing values for most recent quarters raises a red flag over quality of sample

```{r}
# calculate total issued by loan status
# var used to calculate metric: 'loan_amt'
df1_total_issued <- lending_club %>%
  part1_total_issued()

head(df1_total_issued, n = 20)
```

## step 2
## principal payments

```{r}
# calculate principal payments received
# var used to calculate metric: 'total_rec_prncp'
df2_principal <- lending_club %>%
  part2_principal_calc()
head(df2_principal, n = 20)
```

## step 3
## interest payments received

```{r}
# calculate interest payments received
# var used to calculate metric: 'total_rec_int'
df3_rec_int <- lending_club %>%
  part3_rec_int()
head(df3_rec_int, n = 20)
```

## step 4
## average interest rates

```{r}
# calculate average interest rate
# var used to calculate metric: 'int_rate_dbl'*
# *numeric version of original character var, 'int_rate'
df4a_avg_rate <- lending_club %>%
  dplyr::group_by(quarter) %>%
  dplyr::summarise(`Avg. Interest Rate` = mean(int_rate_dbl)) %>%
  dplyr::mutate(grade = "All")
  
df4b_avg_rate <- lending_club %>%
  dplyr::group_by(quarter, grade) %>%
  dplyr::summarise(`Avg. Interest Rate` = mean(int_rate_dbl)) 
  
df4_avg_rate <- df4a_avg_rate %>%
  bind_rows(df4b_avg_rate) %>%
  dplyr::mutate(`Avg. Interest Rate` = percentLabel(`Avg. Interest Rate`*100),
                grade = factor(grade, 
                               levels = c("A", "B", "C", "D", "E", "FG", "All"))) %>%
  dplyr::arrange(quarter)
```

## step 5
## adj. net annualized return

```{r}
# calculate adj. net annualized return
```

## step 6
## join all summary tables by quarter and grade

```{r}
# merge all summary tables into single table
table1 <- df1_total_issued %>%
  dplyr::left_join(df2_principal, by = c("quarter", "grade")) %>%
  dplyr::left_join(df3_rec_int, by = c("quarter", "grade")) %>%
  dplyr::left_join(df4_avg_rate, by = c("quarter", "grade")) %>%
  dplyr::rename(Quarter = quarter,
                Grade = grade)

write.csv(table1, glue::glue("output/lendingclub_table_clean_{month_day}.csv"))

head(table1)
```

```{r}
# create table for display
flextable1 <- flextable(table1)
flextable1
```






